Asssesment of zonal symmetric and asymmetric components of the Southern Annular Mode using a novel approach

### 2020-10-09 analysis/paper/paper.Rmd
Agrega comentarios leandro

---
twocol: FALSE  #  Use FALSE for submission. 
title: Asssesment of zonal symmetric and asymmetric components of the Southern Annular Mode using a novel approach
author1: 
  name: Elio Campitelli
  email: elio.campitelli@cima.fcen.uba.ar
author2: 
  name: Leandro B. Díaz
affiliation: |
  Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
  CONICET – Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
  CNRS – IRD – CONICET – UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina
extra_authors: 
  - name: Carolina Vera
abstract: |
  Enter the text of your abstract here. This is a sample American Meteorological Society (AMS) \LaTeX\ template.  This document provides authors with instructions on the use of the AMS \LaTeX\ template.  Authors should refer to the file amspaper.tex to review the actual \LaTeX\ code used to create this document. The template.tex file should be modified by authors for their own manuscript.
statement: |
  This is significant becasue I wrote it.
bibliography: "AsymSAM.bib,packages.bib"
output: 
  rticles::ams_article:
    latex_engine: "xelatex"
header-includes: 
 - \usepackage{gensymb}
 - \usepackage{subfig}
 - \usepackage[inline]{showlabels}
---

@@

The Southern Annular Mode (SAM) is the main mode of variability in the Southern Hemisphere extratropical circulation [@rogers1982] in daily, monthly, and decadal timescales [@baldwin2001a;@fogt2006] and exerts an important influence in weather conditions such as temperature and precipitation anomalies and sea ice concentration [@fogt2020]. Its positive phase is traditionally described as anomalously low pressures over Antarctica surrounded by a ring of anomalous high pressures in middle-to-high latitudes.

@@

Even though these indices are based on zonal averages, the spatial the spatial structure of the SAM computed from them contains noticeable deviations from zonal symmetry, particularly in the Pacific Ocean region. These zonal asymmetries have not been widely studied, but previous work suggest that they strongly modulate the regional impacts of the SAM, going as far as reversing its relationship between precipitation in South America [@silvestri2009]. At the very least, the fact that the SAM is not entirely zonally symmetric hinders our ability to reconstruct its historical variability prior to the availability of dense observations in the Southern Hemisphere [@jones2009].

At least some of the variability associated with the zonal asymmetries of the SAM is probably forced by the tropics. In particular, ENSO-like variability affects the Southern Hemisphere extratopics through the Pacific-South American Pattern [@mo1987; @kidson1988;@karoly1989], whose wave train protects strongly onto the zonal anomalies corresponding to the SAM in the Pacific sector. And although the relationship between ENSO and SAM is far from simple, tropical influences on the SAM have been observed [@fan2007; @fogt2011; @clem2013]. In particular, @fan2007 computed SAM indices of the Western and the Eastern Hemisphere separately and found that they were much more correlated if the (linear) signal of the ENSO was removed. While this relates to temporally coherent variability of the two hemisphere and not necessarily to zonal asymmetries in the associated spatial patterns, it is nonetheless consistent of the ENSO playing a crucial role in zonal asymmetries of the SAM.

Positive trends in SAM index have been documented by various researchers using different indices mostly on boreal Summer and Autumn [@fogt2020 and references therein]. These trends are thought of driven primarily by stratospheric ozone depletion and understood in the context of zonal mean variables. However, it's not clear how or if the asymmetric component responds to this forcing, or whether its variability could be masking influencing the observed trends independently.

Similarly unclear are the specific impacts of the zonally asymmetric component of the SAM. Positive phase of the SAM is associated with generally colder temperatures over Antarctica and warmer temperatures at higher latitudes [@jones2019] (and vice versa for negative SAM), but there are significant deviations from this zonal mean response, notably in the Antarctic Peninsula and the South Atlantic [@fogt2012]. The SAM signal in precipitation behaves similarly, although with even greater deviation from zonal symmetry [@lim2016]. The importance of zonal asymmetries of the SAM in these impacts have been studied in certain regions. For example, the SAM-precipitation relationship in Southeastern South America and Southern Brazil can be explained by the PSA-like zonally asymmetric circulation associated with the SAM [@silvestri2009; @rosso2018]. @fan2007 also found that precipitation in East Asia was impacted by the variability of only the Western Hemisphere part of the SAM.

We are not aware of any previous work which quantifies the temporal variability of the asymmetric component of the SAM with the exception of @fogt2012. However, their methods based on composites of positive and negative SAM events leads to some issues, such as spatial patterns derived from as little as 4 cases and from imbalanced periods (for example, 5 of the 7 cases in their DJF SAM+ composite are from later than 1988, whereas all of the 8 years in their DJF SAM- composite are from earlier than 1988). This is particularly important due to the inhomogeneities in reanalysis products prior to the satellite era and the possible change in the asymmetric structure of the SAM [@silvestri2009].

Moreover, @fogt2012 studied the zonal asymmetric component of the SAM only in sea level pressure. Zonal asymmetries in the SAM spatial pattern are fairly barotropic throughout the troposphere, but they change dramatically in the stratosphere [@baldwin2009].

@@

@@

To describe the relationship between the SAM indices and temperature and precipitation, we use temperature data from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 [@vose2012; @smith2008], which blends land and ocean temperature analysis into a monthly global grid 5\degree\ longitude by 5\degree\ latitude, and monthly rainfall 0.5\degree\ longitude by 0.5\degree\ latitude data from the Global Precipitation Climatology Centre [@schneider2015].

@@

Our method assumes linearity in the asymmetric component of the SAM. That is, we assume that zonal symmetries associated with positive SAM are almost opposite and of the same magnitude to the ones associated with negative SAM. @fogt2012's composites (their Figure 4) suggest that this might not be entirely valid, although we argue that much of that apparent non-linearity is due to the heterogenous nature of the selected years for constructing the composites. Using our data (from 1979 to 2018), seasonal composites of zonal anomalies of 700 hPa geopotential height for SAM+ (Full SAM index greater than 1 standard deviation) and SAM- (less than minus 1 standard deviation) show relatively high pattern correlations for all seasons and are visually very linear (Figure A9). Therefore, we believe that our method is a reasonable approximation of the phenomenon.

By computing a single EOF pattern using data for all months we are assuming that the zonal anomalies of the SAM are the same in all seasons --December to February (DJF), March to May (MAM), June to August (JJA) and September to November (SON). Geopotential zonal anomalies computed by projecting the first EOF *of each season* are very similar to each other (Figure A10) and show pattern correlations between 0.65 (DJF with JJA) and 0.9 (between MAM and SON). Based on this, we believe that our initial assumption is not unreasonable.

@@

where $\lambda$ and $\phi$ are the longitude and latitude, $t$ is the time, $\alpha$ and $\beta$ are the linear regression coefficients, $X_0$ and $\epsilon$ are the constant and error terms. From this equation, $\alpha$ represents the (linear) association of $X$ with the variability of the Asymmetric SAM that is not explained by the variability of the Symmetric SAM; i.e. it is proportional to the partial correlation of $X$ and the Asymmetric SAM, controlling for the effect of the Symmetric SAM and vice versa for $\beta$. When performing a separate regression for each trimester (DJF, MAM, JJA, SON), we average of the relevant variables seasonally for each year and trimester before computing the regression.

@@

We computed linear trends by Ordinary Least Squares and the 95% confidence interval assuming a t-distribution of the appropriate residual degrees of freedom.

To compute the amplitude of the zonal waves we computed the Fourier transform of the spatial field at each latitude circle.

@@

@@

@@

We can see that the amplitude and latitudinal distribution zonal waves in the Asymmetric SAM on one hand, and the zonal mean in the Symmetric SAM on the other correspond almost exactly to the amplitude and latitudinal distribution in the Full SAM. This confirms the correct decomposition of the SAM in its symmetric and asymmetric components.

Looking at panel b.2 from Figure \ref{fig:2d-regr}, it becomes apparent that zonal waves 1 and 2 modulate the amplitude of zonal wave 3, which --as mentioned before-- is larger in the Western Hemisphere than in the Eastern Hemisphere.

@@

The wave-3 pattern from Figure \ref{fig:2d-regr} panel b.2 is very similar to the Pacific-South American Pattern [@mo1987; @kidson1988] which is a teleconnection pattern associated with the ENSO [@karoly1989]. Indeed, @fogt2011 showed that there is a significant relationship between the SAM and the ENSO. The correlation between the Full SAM and the ENSO as measured by the Oceanic Niño Index [@bamston1997] (ONI) is `r cor_texts[term == "full"]$cor` (p-value = `r cor_texts[term == "full"]$p.value`). Consistent with @fan2007, we show that this relationship is captured entirely by the Asymmetric SAM, as this index has a partial correlation of `r enso_sam_partial[index == "oni" & term == "asym"]$partial_correlation` (p-value = `r enso_sam_partial[index == "oni" & term == "asym"]$p.value`) with the ONI controlling for the effect of the Symmetric SAM, whereas the Symmetric SAM's partial correlation with the ONI is essentially null (`r enso_sam_partial[index == "oni" & term == "sym"]$partial_correlation`; p-value = `r enso_sam_partial[index == "oni" & term == "sym"]$p.value`). We performed the same analysis using the Multivariate ENSO Index [@wolter2011] and the Southern Oscillation Index [@ropelewski1987] to conclude that these results do not depend on the ENSO index used.

## Impacts
\label{sec:impacts}

@@

@@

@@

In Australia the Full SAM is associated with positive precipitation anomalies in South East and this is explained by the Symmetric SAM. However, the Asymmetric SAM is associated with a small area of positive precipitation anomalies in the Eastern Coast of West Australia, maybe due to advection of moist air from the Indian Ocean.In South America, precipitation anomalies associated with the Full SAM are negative both in Southern Chile and Southeastern South America, and positive in Southern Brazil. This features are cleanly separated between the Asymmetric and Symmetric components. The Symmetric SAM explains the negative anomalies in Southern Chile and the Asymmetric SAM, the negative-positive dipole between Southeastern South America and Southern Brazil. Individual seasons mostly follow this pattern.

>>>>

---
twocol: FALSE  #  Use FALSE for submission. 
title: Asssesment of zonal symmetric and asymmetric components of the Southern Annular Mode using a novel approach
author1: 
  name: Elio Campitelli
  email: elio.campitelli@cima.fcen.uba.ar
author2: 
  name: Leandro B. Díaz
affiliation: |
  Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
  CONICET – Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
  CNRS – IRD – CONICET – UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina
extra_authors: 
  - name: Carolina Vera
  # LEAN: A Carolina la podes agrgear como author3 con nuestra misma afiliación 
abstract: |
  Enter the text of your abstract here. This is a sample American Meteorological Society (AMS) \LaTeX\ template.  This document provides authors with instructions on the use of the AMS \LaTeX\ template.  Authors should refer to the file amspaper.tex to review the actual \LaTeX\ code used to create this document. The template.tex file should be modified by authors for their own manuscript.
statement: |
  This is significant becasue I wrote it.
bibliography: "AsymSAM.bib,packages.bib"
output: 
  rticles::ams_article:
    latex_engine: "xelatex"
header-includes: 
 - \usepackage{gensymb}
 - \usepackage{subfig}
 - \usepackage[inline]{showlabels}
---

@@

The Southern Annular Mode (SAM) is the main mode of variability in the Southern Hemisphere extratropical circulation [@rogers1982] in daily, monthly, and decadal timescales [@baldwin2001a;@fogt2006] and exerts an important influence in weather conditions such as temperature and precipitation anomalies and sea ice concentration [e.g. @fogt2020]. Its positive phase is traditionally described as anomalously low pressures over Antarctica surrounded by a ring of anomalous high pressures in middle-to-high latitudes.

@@

Even though these indices are based on zonal averages, the spatial structure of the SAM computed from them contains noticeable deviations from zonal symmetry, particularly in the Pacific Ocean region. These zonal asymmetries have not been widely studied, but previous work suggest that they strongly modulate the regional impacts of the SAM, going as far as reversing its relationship between precipitation in South America [@silvestri2009]. At the very least, the fact that the SAM is not entirely zonally symmetric hinders our ability to reconstruct its historical variability prior to the availability of dense observations in the Southern Hemisphere [@jones2009].

#LEAN: En el "from them" no me queda claro exactamente a qué se refiere el them, me parece que podrías ser un poco más especifíco.

#LEAN:  Creo que deberían agregarse algunas otras referencias para "previous work suggest that they strongly modulate the regional impacts of the SAM"

At least some of the variability associated with the zonal asymmetries of the SAM is probably forced by the tropics. In particular, ENSO-like variability affects the Southern Hemisphere extratopics through the Pacific-South American Pattern [@mo1987; @kidson1988;@karoly1989], whose wave train projects strongly onto the zonal anomalies corresponding to the SAM in the Pacific sector. And although the relationship between ENSO and SAM is far from simple, tropical influences on the SAM have been observed [@fan2007; @fogt2011; @clem2013]. In particular, @fan2007 computed SAM indices of the Western and the Eastern Hemisphere separately and found that they were much more correlated if the (linear) signal of the ENSO was removed. While this relates to temporally coherent variability of the two hemisphere and not necessarily to zonal asymmetries in the associated spatial patterns, it is nonetheless consistent of the ENSO playing a crucial role in zonal asymmetries of the SAM.

Positive trends in SAM index have been documented by various researchers using different indices mostly on boreal Summer and Autumn [e.g. @fogt2020 and references therein]. These trends are thought of driven primarily by stratospheric ozone depletion and understood in the context of zonal mean variables. However, it's not clear how or if the asymmetric component responds to this forcing, or whether its variability could be masking influencing the observed trends independently.

#LEAN: Te faltaría alguna referencia de la segunda oración "These trends are thought of driven primarily by stratospheric ozone depletion and understood in the context of zonal mean variables.". Se me ocurre alguno de Julie arblaster que tiene varios trabajos en esa línea. También agregaría que los GHG también tienen efecto.

Similarly unclear are the specific impacts of the zonally asymmetric component of the SAM. Positive phase of the SAM is associated with generally colder temperatures over Antarctica and warmer temperatures at higher latitudes [@jones2019] (and vice versa for negative SAM), but there are significant deviations from this zonal mean response, notably in the Antarctic Peninsula and the South Atlantic [@fogt2012]. The SAM signal in precipitation behaves similarly, although with even greater deviation from zonal symmetry [@lim2016]. The importance of zonal asymmetries of the SAM on these impacts have been studied in certain regions. For example, the SAM-precipitation relationship in Southeastern South America and Southern Brazil can be explained by the PSA-like zonally asymmetric circulation associated with the SAM [@silvestri2009; @rosso2018]. @fan2007 also found that precipitation in East Asia was impacted by the variability of only the Western Hemisphere part of the SAM.

We are not aware of any previous work which quantifies the temporal variability of the asymmetric component of the SAM with the exception of @fogt2012. However, their methods based on composites of positive and negative SAM events leads to some issues, such as spatial patterns derived from as little as 4 cases and from imbalanced periods (for example, 5 of the 7 cases in their DJF SAM+ composite are from later than 1988, whereas all of the 8 years in their DJF SAM- composite are from earlier than 1988). This is particularly important due to the inhomogeneities in reanalysis products prior to the satellite era and the possible change in the asymmetric structure of the SAM [@silvestri2009]. Moreover, @fogt2012 studied the zonal asymmetric component of the SAM only in sea level pressure. Zonal asymmetries in the SAM spatial pattern are fairly barotropic throughout the troposphere, but they change dramatically in the stratosphere [@baldwin2009].

@@

#LEAN: Te había hecho el comentario antes pero quizás no había sido del todo claro. Me parece que acá hay que distinguir más en cómo se trabaja metodológicamente cada parte con respecto a los niveles atmosféricos. Por ej. para la evaluación de los impactos solamente se está considerando el ínidice troposférico.

@@

To describe the relationship between the SAM indices and temperature and precipitation, we use temperature data from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 [@smith2008; @vose2012], which blends land and ocean temperature analysis into a monthly global grid 5\degree\ longitude by 5\degree\ latitude, and monthly rainfall 0.5\degree\ longitude by 0.5\degree\ latitude data from the Global Precipitation Climatology Centre [@schneider2015]. The rainfall product uses station-based records, and thus it only has continental coverage.

#LEAN: Se suele citar esto tambień para GPCC, sería el paper correspondiente de la base https://doi.org/10.3390/atmos8030052

@@

#LEAN: Hay algún problema con los guiones en los figure captions, revisa que en varios el PDF quedo todo pegado 19792018

Our method assumes linearity in the asymmetric component of the SAM. That means we assume that zonal symmetries associated with positive SAM are almost opposite and of the same magnitude to the ones associated with negative SAM. @fogt2012's composites (their Figure 4) suggest that this might not be entirely valid, although we argue that much of that apparent non-linearity is due to the heterogenous nature of the selected years for constructing the composites. Using our data (from 1979 to 2018), seasonal composites of zonal anomalies of 700 hPa geopotential height for SAM+ (Full SAM index greater than 1 standard deviation) and SAM- (less than minus 1 standard deviation) show relatively high pattern correlations for all seasons and are visually very linear (Figure A9). Therefore, we believe that our method is a reasonable approximation of the phenomenon.

#Para la Figure A9, A10 también aclara el período

#Un detalle que habría que aclarar mejor en algún lado es por qué utilizas 700 hPa para esta discusión y otras que siguen. Sin la aclaración puede zonar un poco arbitrario (y acá solo consideras ese nivel para la justificación metodológica).

By computing a single EOF pattern using data for all months we are assuming that the zonal anomalies of the SAM are the same in all seasons --December to February (DJF), March to May (MAM), June to August (JJA) and September to November (SON). Geopotential zonal anomalies computed by projecting the first EOF of each season independently are very similar to each other (Figure A10) and show pattern correlations between 0.65 (DJF with JJA) and 0.9 (between MAM and SON). Based on this, we believe that our initial assumption is not unreasonable.

@@

where $\lambda$ and $\phi$ are the longitude and latitude, $t$ is the time, $\alpha$ and $\beta$ are the linear regression coefficients, $X_0$ and $\epsilon$ are the constant and error terms. From this equation, $\alpha$ represents the (linear) association of $X$ with the variability of the Asymmetric SAM that is not explained by the variability of the Symmetric SAM; i.e. it is proportional to the partial correlation of $X$ and the Asymmetric SAM, controlling for the effect of the Symmetric SAM and vice versa for $\beta$. When performing a separate regression for each trimester (DJF, MAM, JJA, SON), we average the relevant variables seasonally for each year and trimester before computing the regression.

@@

We computed linear trends by Ordinary Least Squares and the 95% confidence interval assuming a t-distribution of the appropriate residual degrees of freedom. To compute the amplitude of the zonal waves we computed the Fourier transform of the spatial field at each latitude circle.

@@

A version-controlled repository of the code used to create this analysis can be found at https://github.com/eliocamp/asymsam.
#LEAN: Ver de mover a dónde debería ir

@@

#LEAN: La aclaración metodológica de los "density estimates" no debería aparecer en el caption sino en la metodología del trabajo. Después en el caption ya no es necesario volverlo a aclarar. Algo que me resulta raro en la figura es la posición de las letras. Creo que es mejor darla vuelta, poner la aclaración de los niveles a la izquierda y los density estimates del lado derecho, eso la va a hacer bastante más clara y también más chorente con el caption.

@@

We can see that the amplitude and latitudinal distribution zonal waves in the Asymmetric SAM on one hand, and the zonal mean in the Symmetric SAM on the other correspond almost exactly to the amplitude and latitudinal distribution in the Full SAM. This confirms the correct decomposition of the SAM in its symmetric and asymmetric components. Looking at panel b.2 from Figure \ref{fig:2d-regr}, it becomes apparent that zonal waves 1 and 2 modulate the amplitude of zonal wave 3, which --as mentioned before-- is larger in the Western Hemisphere than in the Eastern Hemisphere.

@@

The wave-3 pattern from Figure \ref{fig:2d-regr} panel b.2 is very similar to the Pacific-South American Pattern [@mo1987; @kidson1988] which is a teleconnection pattern associated with the ENSO [@karoly1989]. Indeed, @fogt2011 showed that there is a significant relationship between the SAM and the ENSO. The correlation between the Full SAM and the ENSO as measured by the Oceanic Niño Index [ONI, @bamston1997] is `r cor_texts[term == "full"]$cor` (p-value = `r cor_texts[term == "full"]$p.value`). Consistent with @fan2007, we show that this relationship is captured entirely by the Asymmetric SAM, as this index has a partial correlation of `r enso_sam_partial[index == "oni" & term == "asym"]$partial_correlation` (p-value = `r enso_sam_partial[index == "oni" & term == "asym"]$p.value`) with the ONI controlling for the effect of the Symmetric SAM, whereas the Symmetric SAM's partial correlation with the ONI is essentially null (`r enso_sam_partial[index == "oni" & term == "sym"]$partial_correlation`; p-value = `r enso_sam_partial[index == "oni" & term == "sym"]$p.value`). We performed the same analysis using the Multivariate ENSO Index [@wolter2011] and the Southern Oscillation Index [@ropelewski1987] to conclude that these results do not depend on the ENSO index used.

## Surface Impacts
\label{sec:impacts}

@@

#LEAN: En las figuras 11 y 12 deberías decir en el caption algo sobre cuál es el intervalo de los contornos de la altura geopotencial

@@

#LEAN: Te quedo distinto el estilo en que se citan  @silvestri2009 y @zamboni2010

@@

In Australia the Full SAM is associated with positive precipitation anomalies in South East and this is explained by the Symmetric SAM. However, the Asymmetric SAM is associated with a small area of positive precipitation anomalies in the Eastern Coast of West Australia, maybe due to advection of moist air from the Indian Ocean. In South America, precipitation anomalies associated with the Full SAM are negative both in Southern Chile and Southeastern South America, and positive in Southern Brazil. This features are cleanly separated between the Asymmetric and Symmetric components. The Symmetric SAM explains the negative anomalies in Southern Chile and the Asymmetric SAM, the negative-positive dipole between Southeastern South America and Southern Brazil. Individual seasons mostly follow this pattern.
